Effectful Semantics for Substructural SSA

### 2023-11-17 pldi24/paper.tex
Tap

\begin{itemize}
\item We give a type system for SSA-style programs, which has a full
substitution theorem and equational theory. This language includes substructural
types, both for controlling effects and tracking ownership in values, and our
strong substitution principle makes formulating inlining and rewriting
optimisations easier than in the traditional CFG-based formulations. 
\item We give an axiomatic semantics for our language, by showing that it has a
compositional intepretation in any effectful category with an Elgot structure.
We use this structure to give generic proofs of the soundness of substitution
and rewriting properties.
\item Our abstract semantics justifies a variety of powerful optimisation
techniques, inlining, loop fusion, hoisting, and strength reduction. 
\item We show how Elgot monads give rise to instances of our categorical
semantics. We use this to derive semantics from a general trace monad, and show
that various monad transformers such as state transformers preserve the needed
Elgot structure. 
\item We use these tools to demonstrate the existence of a variety of concrete
models satisfying these categorical axioms. Starting first from simple languges
such as state plus printing, we also show how more challenging semantics such as
the TSO semantics for weak memory also fit into our framework. This thus gives
rise to an SSA-based IR with support for weak memory operations, which is fully
semantically-justified.
% \item Our weak memory semantics and our results about Elgot monads are
% formalised in the Lean theorem prover.
\end{itemize}

@@

We parametrize an \isotopessa dialect by a set of \textit{base types} \(X \in
\mc{T}\) and, for each pair of types \(A, B\), a set of \textit{instructions}
\(f \in \mc{I}^q_p\) with \textit{centrality} \(p \in \{0, 1\}\) and
\textit{quantity} \(q \subseteq \tint\). As described above, a \textit{central}
instruction is one for which execution order does not matter, while the quantity
\(q\) indicates whether it is safe to delete an instruction if its result is
unused (in which case the instruction is \textit{affine}, and \(\taff \subseteq
q\)) or duplicate an instruction used multiple times (in which case the
instruction is \textit{relevant}, and \(\trel \subseteq q\)). An instruction
which is both affine and relevant is called \textit{pure}. In general, the set
of types is generated by base types \(X\), the \textit{unit type} \(\mb{1}\),
the \textit{boolean type} \(\mb{2}\), and tensor products of types \(A \otimes
B\). We write \(\otimes\) to underscore the fact that types are allowed to be
substructural: every type is assigned a \textit{linearity} \(\ms{lin}(A)
\subseteq \tint\) satisfying the constraints in Figure \ref{fig:ssa-structural}.
In particular, the rules \rle{unit-lin} and \rle{bool-lin} state that both the
unit and boolean types must be intuitionistic; the rule \rle{pair-lin} says that
the linearity of a pair is simply the intersection of the linearity of its
components. 
% The rules \rle{nil-lin} and \rle{cons-lin} together say that the
% linearity of a context is the intersection of the linearity of its variables,
% where the linearity of a variable is the intersection of the linearity of its
% type and its quantity annotation.

@@

We begin with some standard definitions:
\begin{definition}[Binoidal Category] 
  A \textbf{binoidal category} \(\mc{C}\) is a category equipped with a binary
  operation on objects \(\otimes: \mc{C} \times \mc{C} \to \mc{C}\) which is
  independently functorial in each argument (i.e., for each object \(A\), \(A
  \otimes -\) and \(- \otimes A\) are functors).
\end{definition}
Given \(f: \mc{C}(A, B)\), \(g: \mc{C}(A', B')\) in a binoidal category, we
write \(f \ltimes g = f \otimes A';B \otimes g\) and \(f \rtimes g = A \otimes
g;f \otimes B'\); we say \(f\) is \textbf{central} if, for all \(g\), \(f
\ltimes g = f \rtimes g\) and \(g \ltimes f = g \rtimes f\), in which case we
simply write \(f \otimes g\), \(g \otimes f\) respectively.
\begin{definition}[Premonoidal Category] 
  A \textbf{premonoidal category} is a binoidal category with an identity object
  \(I: \mc{C}\) such that there exist central natural isomorphisms \(\lambda_A:
  I \otimes A \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I
  \simeq A\) (the \textbf{right unitor}), and \(\alpha_{A, B, C}: (A \otimes B)
  \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
  satisfying the \textit{pentagon} and \textit{triangle laws} (see e.g.
  \citet{maclane:71}). We say a premonoidal category is \textbf{symmetric} if it
  is equipped with a central natural isomorphism \(\sigma_{A, B}: A \otimes B
  \simeq B \otimes A\) (the \textbf{symmetry}) satisfying the \textit{hexagon
  law} (see \cite{maclane:71}). We say a premonoidal category is
  \textbf{monoidal} if every morphism is central.
\end{definition}
One issue with the notion of a premonoidal category is that it is unclear what
exactly a premonoidal functor should be: if we want them to compose, we would
like to preserve central morphisms, but trying to preserve \textit{all} central
morphisms is too strict a notion. We want, in essence, to be able to distinguish
morphisms that are truly "pure" from those which simply happen to be central. To
do that, we introduce the notion of an \textit{effectful category} as follows:
\begin{definition}[Effectful Category] 
  An \textbf{effectful category} \(\mc{V} \to \mc{C}\) is an identity on objects
  functor \(\upg{\cdot}{}: \mc{V} \to \mc{C}\) from a monoidal category
  \(\mc{V}\) to a premonoidal category \(\mc{C}\) which preserves all
  premonoidal structure (i.e. sends associators to associators, unitors to
  unitors, and tensor products to tensor products). It is \textbf{symmetric} if
  both \(\mc{V}\) and \(\mc{C}\) are, and \(\upg{\cdot}{}\) preserves the
  symmetry. A \textbf{Freyd category} is an effectful category where \(\mc{V}\)
  is Cartesian.
\end{definition}
Unfortunately, the notion of "purity" given by an effectful category only allows
us to perform relatively weak optimistions in the presence of effects. To remedy
this, we want to be able to make a finer distinction between morphisms which are
merely \textit{central} and those which are also \textit{relevant} (i.e.,
copying the input and applying them twice is the same as applying them once and
copying the output) and/or \textit{affine} (i.e., applying them and discarding
the result is the same as never doing them at all). Doing so also allows us to
add the capability of \textit{distinguishing} between objects which are affine,
relevant, neither or both, enabling us to reason about substructural type
systems categorically.

To do this, we begin by introducing the notion of a \textit{symmetric
premonoidal subcategory}, as follows:
\begin{definition}[(Symmetric) Premonoidal Subcategory]
  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory
  \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is
  closed under whiskering and contains all associators, unitors, and symmetries.
  We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\),
  \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is
  \textbf{central} if this generalizes to the case where only one of \(f, g\) is
  in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).
\end{definition}

@@

    symmetric premonoidal subcategories \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\).
    We define, for \(q \subseteq \{\ms{a}, \ms{r}\}\), \(\mc{C}^q = \bigcap_{\ell \in q}\mc{C}^\ell\), where \(\mc{C}^\varnothing = \mc{C}\)

@@

      \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\ms{a}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)

@@

      \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)

@@

We can now give a semantics for an \isotopessa dialect with types
\(\mc{T}\) and instructions \(\mc{I}\). We'll need:
\begin{itemize}
  \item A \textit{substructural effectful category} \(\mc{C} = \mc{C}_1 \to
  \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot},
  equipped with
  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects
  in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies \ms{base}(X) \in
  \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies \ms{base}(X) \in
  \ms{Rel}(\mc{C})\)
  \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\)
  to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
\end{itemize}

We begin by giving a semantics for types and contexts in terms of objects in
\(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}: we map tensor products to
tensor products, the unit type to the monoidal unit, and booleans to \(\mb{1} +
\mb{1}\). Contexts are simply interpreted as the tensor products of the variable
types which appear in them; in particular, the quantity annotations have
\textit{no} effect on a context's semantics. We then give a semantics for
structural judgements in Figure \ref{fig:ssa-structural-semantics}. In
particular, we begin by interpreting the judgements that a type is affine (i.e.
\(\tlin{A}{\taff}\)) or relevant (i.e. \(\tlin{A}{\trel}\)) as the drop morphism
\(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, I)\) and split morphism
\(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, \dnt{A} \otimes \dnt{A})\) respectively.
We can then interpret a context split \(\csplits{\Gamma}{\Delta}{\Xi}\) as a
pure morphism \(\mc{C}^\tint_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})\)
which partitions the variables as appropriate, splitting and dropping as
necessary. We interpret a context weakening \(\cwk{\Gamma}{\Delta}\) as the
associated split \(\csplits{\Gamma}{\Delta}{\cdot}\) postcomposed with a unitor.
Finally, we interpret a label-set weakening \(\lwk{\ms{L}}{\ms{K}}\) as a pure
morphism \(\mc{C}^\tint_1(\dnt{\ms{L}}, \dnt{\ms{K}})\), where \rle{join-nil} is
the identity, \rle{join-cons} is the sum of \(\dnt{\lwk{\ms{L}}{\ms{K}}}\) and
\(\dnt{\cwk{\Gamma}{\Delta}}\), and \rle{join-zero} is
\(\dnt{\lwk{\ms{L}}{\ms{K}}}\) postcomposed with the left injection.

\begin{figure}
  \begin{gather*}
    \boxed{\dnt{A}}: |\mc{C}| \\
    \dnt{X} = \ms{base}(X) 
    \qquad
    \dnt{\mb{1}} = \mb{1}
    \qquad
    \dnt{\mb{2}} = \mb{2} = \mb{1} + \mb{1}
    \qquad \dnt{A \otimes B} = \dnt{A} \otimes \dnt{B} \\
    \boxed{\dnt{\Gamma}}: |\mc{C}| \\
    \dnt{\cdot} = \mb{1} \qquad 
    \dnt{\Gamma, \thyp{x}{A}{q}} = \dnt{\Gamma} \otimes \dnt{A} \\
    \boxed{\dnt{\ms{L}}}: |\mc{C}| \\
    \dnt{\cdot} = \mb{0} \qquad
    \dnt{\ms{L}, \llhyp{\ell}{\Gamma}{A}} = 
      \dnt{\ms{L}} + \dnt{\Gamma} \otimes \dnt{A}
  \end{gather*}
  \caption{Semantics for \isotopessa types and contexts}
  \Description{Semantics for isotope-SSA types and contexts}
  \label{fig:ssa-type-semantics}
\end{figure}

\begin{figure}
  \begin{gather*}
    \boxed{\dnt{\tlin{A}{\taff}: \mc{C}_1^\tint(\dnt{A}, \mb{1})}}
      \\
      \dnt{\tlin{A}{\taff}} = \ms{drop}(\dnt{A})
      % \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
      % \dnt{\tlin{A \otimes B}{\taff}} 
      %   = \dnt{\tlin{A}{\taff}} \otimes \dnt{\tlin{B}{\taff}}
      %   ; \lambda 
      \\
    \boxed{\dnt{\tlin{A}{\trel}}
      : \mc{C}_1^\tint(\dnt{A}, \dnt{A} \otimes \dnt{A})}
      \\
      \dnt{\tlin{A}{\trel}} = \ms{split}(\dnt{A})
      % \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
      % \dnt{\tlin{A \otimes B}{\trel}} = 
      %   \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
      %   ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha 
      \\
    \boxed{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})}
      \\
    \dnt{\csplits{\cdot}{\cdot}{\cdot}} = \lambda^{-1} \qquad
    \dnt{\csplits
      {\Gamma, \thyp{x}{A}{q}}
      {\Delta, \thyp{x}{A}{q}}
      {\Xi}} 
      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha;\dnt{\Delta} \otimes \sigma;\alpha 
      \\
    \dnt{\csplits
      {\Gamma, \thyp{x}{A}{q}}
      {\Delta}
      {\Xi, \thyp{x}{A}{q}}} 
      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha 
      \\
    \dnt{\csplits
      {\Gamma, \thyp{x}{A}{q}}
      {\Delta}
      {\Xi, \thyp{x}{A}{q}}} \
      = \dnt{\Gamma} \otimes \dnt{\tlin{A}{\trel}};
        \alpha;
        \dnt{\Delta} \otimes \sigma \otimes \dnt{A};
        \alpha 
        \\
    \dnt{\csplits
      {\Gamma, \thyp{x}{A}{q}}
      {\Delta}
      {\Xi}}
      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
        \otimes \dnt{\tlin{A}{\taff}}
      ; \rho 
      \\
    \boxed{\dnt{\cwk{\Gamma}{\Delta}}
      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta})} 
      \\
      \dnt{\cwk{\Gamma}{\Delta}} 
      = \dnt{\csplits{\Gamma}{\Delta}{\cdot}};\rho 
      \\
    \boxed{\dnt{\lwk{\ms{L}}{\ms{K}}}
      : \mc{C}_1^\tint(\dnt{\ms{L}}, \dnt{\ms{K}})} 
      \\
    \dnt{\lwk{\cdot}{\cdot}} = \ms{id} \qquad
    \dnt{\lwk
      {\ms{L}, \llhyp{\ell}{\Gamma}{A}}
      {\ms{K}, \llhyp{\ell}{\Delta}{A}}}
      = \dnt{\lwk{\ms{L}}{\ms{K}}} + \upg{(\dnt{\cwk{\Gamma}{\Delta}} \otimes \dnt{A})}{} 
      \\
    \dnt{\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      = \dnt{\lwk{\ms{L}}{\ms{K}}};\iota_0
  \end{gather*}
  \caption{Semantics for \isotopessa structural judgements}
  \Description{Semantics for isotope-SSA structural judgements}
  \label{fig:ssa-structural-semantics}
\end{figure}

@@

Because we have a type-theoretic presentation of control-flow graphs, we also
have a substitution principle and equational theory for them. This lets us show
the soundness of inlining, loop unrolling, loop fusion, dead-code elimination,
and other optimisations which change the structure of the control-flow graph.

Note that we have not yet proved anything about the dataflow analyses used to
find the rewrites -- we have only shown that the rewrites themselves are
semantically justified.

>>>>

\begin{itemize}
\item We give a type system for SSA-style programs, which has a full
substitution theorem and equational theory. This language includes substructural
types, both for controlling effects and tracking ownership in values, and our
strong substitution principle makes formulating inlining and rewriting
optimisations easier than in the traditional CFG-based formulations. 
\item We give an axiomatic semantics for our language, by showing that it has a
compositional intepretation in any effectful category with an Elgot structure.
We use this structure to give generic proofs of the soundness of substitution
and rewriting properties.
\item Our abstract semantics justifies a variety of powerful optimisation
techniques, including inlining, loop fusion, hoisting, and strength reduction. 
\item We show how Elgot monads give rise to instances of our categorical
semantics. We use this to derive semantics from a general trace monad, and show
that various monad transformers such as state transformers preserve the needed
Elgot structure. 
\item We use these tools to demonstrate the existence of a variety of concrete
models satisfying these categorical axioms. Starting first from simple languges
such as state plus printing, we also show how more challenging semantics such as
the TSO semantics for weak memory also fit into our framework. This thus gives
rise to an SSA-based IR with support for weak memory operations, which is fully
semantically-justified.
% \item Our weak memory semantics and our results about Elgot monads are
% formalised in the Lean theorem prover.
\end{itemize}

@@

We parametrize our syntax by a set of \textit{base types} \(X \in \mc{T}\) and,
for each pair of types \(A, B\), a set of \textit{instructions} \(f \in
\mc{I}^q_p\) with \textit{centrality} \(p \in \{0, 1\}\) and \textit{quantity}
\(q \subseteq \tint\). As described above, a \textit{central} instruction is one
for which execution order does not matter, while the quantity \(q\) indicates
whether it is safe to delete an instruction if its result is unused (in which
case the instruction is \textit{affine}, and \(\taff \subseteq q\)) or duplicate
an instruction used multiple times (in which case the instruction is
\textit{relevant}, and \(\trel \subseteq q\)). An instruction which is both
affine and relevant is called \textit{pure}. The set of types is generated by
base types \(X\), the \textit{unit type} \(\mb{1}\), the \textit{boolean type}
\(\mb{2}\), and tensor products of types \(A \otimes B\). We write \(\otimes\)
to underscore the fact that types are allowed to be substructural: similarly to
instructions, every type is assigned a \textit{linearity} \(\ms{lin}(A)
\subseteq \tint\) satisfying the constraints in Figure \ref{fig:ssa-structural},
except that a type which is both affine and relevant is called
\textit{intuitionistic}. In particular, the rules \rle{unit-lin} and
\rle{bool-lin} state that both the unit and boolean types must be
intuitionistic; the rule \rle{pair-lin} says that the linearity of a pair is
simply the intersection of the linearity of its components. 
% The rules \rle{nil-lin} and \rle{cons-lin} together say that the
% linearity of a context is the intersection of the linearity of its variables,
% where the linearity of a variable is the intersection of the linearity of its
% type and its quantity annotation.

@@

We begin with some standard definitions:
\begin{definition}[Binoidal Category] 
  A \textbf{binoidal category} \(\mc{C}\) is a category equipped with a binary
  operation on objects \(\otimes: \mc{C} \times \mc{C} \to \mc{C}\) which is
  independently functorial in each argument (i.e., for each object \(A\), \(A
  \otimes -\) and \(- \otimes A\) are functors).
Given \(f: \mc{C}(A, B)\), \(g: \mc{C}(A', B')\) in a binoidal category, we
write \(f \ltimes g = f \otimes A';B \otimes g\) and \(f \rtimes g = A \otimes
g;f \otimes B'\); we say \(f\) is \textbf{central} if, for all \(g\), \(f
\ltimes g = f \rtimes g\) and \(g \ltimes f = g \rtimes f\), in which case we
simply write \(f \otimes g\), \(g \otimes f\) respectively.
\begin{definition}[Premonoidal Category] 
  A \textbf{premonoidal category} is a binoidal category with an identity object
  \(I: \mc{C}\) such that there exist central natural isomorphisms \(\lambda_A:
  I \otimes A \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I
  \simeq A\) (the \textbf{right unitor}), and \(\alpha_{A, B, C}: (A \otimes B)
  \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
  satisfying the \textit{pentagon} and \textit{triangle laws} (see e.g.
  \citet{maclane:71}). We say a premonoidal category is \textbf{symmetric} if it
  is equipped with a central natural isomorphism \(\sigma_{A, B}: A \otimes B
  \simeq B \otimes A\) (the \textbf{symmetry}) satisfying the \textit{hexagon
  law} (see \cite{maclane:71}). We say a premonoidal category is
  \textbf{monoidal} if every morphism is central.
One issue with the notion of a premonoidal category is that it is unclear what
exactly a premonoidal functor should be: if we want them to compose, we would
like to preserve central morphisms, but trying to preserve \textit{all} central
morphisms is too strict a notion. We want, in essence, to be able to distinguish
morphisms that are truly "pure" from those which simply happen to be central. To
do that, we introduce the notion of an \textit{effectful category} as follows:
\begin{definition}[Effectful Category] 
  An \textbf{effectful category} \(\mc{V} \to \mc{C}\) is an identity on objects
  functor \(\upg{\cdot}{}: \mc{V} \to \mc{C}\) from a monoidal category
  \(\mc{V}\) to a premonoidal category \(\mc{C}\) which preserves all
  premonoidal structure (i.e. sends associators to associators, unitors to
  unitors, and tensor products to tensor products). It is \textbf{symmetric} if
  both \(\mc{V}\) and \(\mc{C}\) are, and \(\upg{\cdot}{}\) preserves the
  symmetry. A \textbf{Freyd category} is an effectful category where \(\mc{V}\)
  is Cartesian.
Unfortunately, the notion of "purity" given by an effectful category only allows
us to perform relatively weak optimistions in the presence of effects. To remedy
this, we want to be able to make a finer distinction between morphisms which are
merely \textit{central} and those which are also \textit{relevant} (i.e.,
copying the input and applying them twice is the same as applying them once and
copying the output) and/or \textit{affine} (i.e., applying them and discarding
the result is the same as never doing them at all). Doing so also allows us to
add the capability of \textit{distinguishing} between objects which are affine,
relevant, neither or both, enabling us to reason about substructural type
systems categorically. To do this, we begin by introducing the notion of a \textit{symmetric
premonoidal subcategory}, as follows:
\begin{definition}[(Symmetric) Premonoidal Subcategory]
  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory
  \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is
  closed under whiskering and contains all associators, unitors, and symmetries.
  We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\),
  \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is
  \textbf{central} if this generalizes to the case where only one of \(f, g\) is
  in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).

@@

    symmetric premonoidal subcategories \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\). We
    define, for \(q \subseteq \{\ms{a}, \ms{r}\}\), \(\mc{C}^q = \bigcap_{\ell
    \in q}\mc{C}^\ell\), where \(\mc{C}^\varnothing = \mc{C}\)

@@

      \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\) satisfying
      \item \(\forall A, B \in \mc{A}. \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\), \(\forall f: \mc{C}^{\ms{a}}(A, B). f;\ms{drop}(B) = \ms{drop}(A)\)

@@

      \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\) satisfying

@@

We can now give a semantics for an \isotopessa dialect with types
\(\mc{T}\) and instructions \(\mc{I}\). We'll need:
\begin{itemize}
  \item A \textit{substructural effectful category} \(\mc{C} = \mc{C}_1 \to
  \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot},
  equipped with
  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects
  in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies \ms{base}(X) \in
  \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies \ms{base}(X) \in
  \ms{Rel}(\mc{C})\)
  \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\)
  to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
\end{itemize}
We begin by giving a semantics for types and contexts in terms of objects in
\(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}: we map tensor products to
tensor products, the unit type to the monoidal unit, and booleans to \(\mb{1} +
\mb{1}\). Contexts are simply interpreted as the tensor products of the variable
types which appear in them; in particular, the quantity annotations have
\textit{no} effect on a context's semantics. We then give a semantics for
structural judgements in Figure \ref{fig:ssa-structural-semantics}. In
particular, we begin by interpreting the judgements that a type is affine (i.e.
\(\tlin{A}{\taff}\)) or relevant (i.e. \(\tlin{A}{\trel}\)) as the drop morphism
\(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, I)\) and split morphism
\(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, \dnt{A} \otimes \dnt{A})\) respectively.
We can then interpret a context split \(\csplits{\Gamma}{\Delta}{\Xi}\) as a
pure morphism \(\mc{C}^\tint_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})\)
which partitions the variables as appropriate, splitting and dropping as
necessary. We interpret a context weakening \(\cwk{\Gamma}{\Delta}\) as the
associated split \(\csplits{\Gamma}{\Delta}{\cdot}\) postcomposed with a unitor.
Finally, we interpret a label-set weakening \(\lwk{\ms{L}}{\ms{K}}\) as a pure
morphism \(\mc{C}^\tint_1(\dnt{\ms{L}}, \dnt{\ms{K}})\), where \rle{join-nil} is
the identity, \rle{join-cons} is the sum of \(\dnt{\lwk{\ms{L}}{\ms{K}}}\) and
\(\dnt{\cwk{\Gamma}{\Delta}}\), and \rle{join-zero} is
\(\dnt{\lwk{\ms{L}}{\ms{K}}}\) postcomposed with the left injection.

\begin{figure}
  \begin{align*}
    \boxed{\dnt{A}: |\mc{C}|} & \qquad  
    \dnt{X} = \ms{base}(X) 
    \qquad
    \dnt{\mb{1}} = \mb{1}
    \qquad
    \dnt{\mb{2}} = \mb{2} = \mb{1} + \mb{1}
    \qquad \dnt{A \otimes B} = \dnt{A} \otimes \dnt{B} \\
    \boxed{\dnt{\Gamma}: |\mc{C}|} & \qquad 
    \dnt{\cdot} = \mb{1} \qquad 
    \dnt{\Gamma, \thyp{x}{A}{q}} = \dnt{\Gamma} \otimes \dnt{A} 
    \\
    \boxed{\dnt{\ms{L}}: |\mc{C}|} & \qquad 
    \dnt{\cdot} = \mb{0} \qquad
    \dnt{\ms{L}, \llhyp{\ell}{\Gamma}{A}} = 
      \dnt{\ms{L}} + \dnt{\Gamma} \otimes \dnt{A}
  \end{align*}
  \caption{Semantics for \isotopessa types and contexts}
  \Description{Semantics for isotope-SSA types and contexts}
  \label{fig:ssa-type-semantics}
\end{figure}

\begin{figure}
  \begin{gather*}
    % \boxed{\dnt{\tlin{A}{\taff}: \mc{C}_1^\tint(\dnt{A}, \mb{1})}}
    %   \\
    %   \dnt{\tlin{A}{\taff}} = \ms{drop}(\dnt{A})
    %   % \dnt{\tlin{X}{\taff}} = \ms{drop}(X) \qquad
    %   % \dnt{\tlin{A \otimes B}{\taff}} 
    %   %   = \dnt{\tlin{A}{\taff}} \otimes \dnt{\tlin{B}{\taff}}
    %   %   ; \lambda 
    %   \\
    % \boxed{\dnt{\tlin{A}{\trel}}
    %   : \mc{C}_1^\tint(\dnt{A}, \dnt{A} \otimes \dnt{A})}
    %   \\
    %   \dnt{\tlin{A}{\trel}} = \ms{split}(\dnt{A})
    %   % \dnt{\tlin{X}{\trel}} = \ms{split}(X) \qquad
    %   % \dnt{\tlin{A \otimes B}{\trel}} = 
    %   %   \dnt{\tlin{A}{\trel}} \otimes \dnt{\tlin{B}{\trel}}
    %   %   ;\alpha;\dnt{A} \otimes \sigma \otimes \dnt{B};\alpha 
    %   \\
    \boxed{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})}
      \\
    \dnt{\csplits{\cdot}{\cdot}{\cdot}} = \lambda^{-1} \qquad
    \dnt{\csplits
      {\Gamma, \thyp{x}{A}{q}}
      {\Delta, \thyp{x}{A}{q}}
      {\Xi}} 
      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha;\dnt{\Delta} \otimes \sigma;\alpha 
      \\
    \dnt{\csplits
      {\Gamma, \thyp{x}{A}{q}}
      {\Delta}
      {\Xi, \thyp{x}{A}{q}}} 
      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}} \otimes \dnt{A};\alpha 
      \\
    \dnt{\csplits
      {\Gamma, \thyp{x}{A}{q}}
      {\Delta}
      {\Xi, \thyp{x}{A}{q}}} \
      = \dnt{\Gamma} \otimes \ms{split}(\dnt{A});
        \alpha;
        \dnt{\Delta} \otimes \sigma \otimes \dnt{A};
        \alpha 
        \\
    \dnt{\csplits
      {\Gamma, \thyp{x}{A}{q}}
      {\Delta}
      {\Xi}}
      = \dnt{\csplits{\Gamma}{\Delta}{\Xi}}
        \otimes \ms{drop}(\dnt{A})
      ; \rho 
      \\
    \boxed{\dnt{\cwk{\Gamma}{\Delta}}
      : \mc{C}_1^\tint(\dnt{\Gamma}, \dnt{\Delta})} 
      \\
      \dnt{\cwk{\Gamma}{\Delta}} 
      = \dnt{\csplits{\Gamma}{\Delta}{\cdot}};\rho 
      \\
    \boxed{\dnt{\lwk{\ms{L}}{\ms{K}}}
      : \mc{C}_1^\tint(\dnt{\ms{L}}, \dnt{\ms{K}})} 
      \\
    \dnt{\lwk{\cdot}{\cdot}} = \ms{id} \qquad
    \dnt{\lwk
      {\ms{L}, \llhyp{\ell}{\Gamma}{A}}
      {\ms{K}, \llhyp{\ell}{\Delta}{A}}}
      = \dnt{\lwk{\ms{L}}{\ms{K}}} + \upg{(\dnt{\cwk{\Gamma}{\Delta}} \otimes \dnt{A})}{} 
      \\
    \dnt{\lwk{\ms{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      = \dnt{\lwk{\ms{L}}{\ms{K}}};\iota_0
  \end{gather*}
  \caption{Semantics for \isotopessa structural judgements}
  \Description{Semantics for isotope-SSA structural judgements}
  \label{fig:ssa-structural-semantics}
\end{figure}

@@

Because we have a type-theoretic presentation of control-flow graphs, we also
have a substitution principle and equational theory for them. This lets us show
the soundness of inlining, loop unrolling, loop fusion, dead-code elimination,
and other optimisations which change the structure of the control-flow graph.
Note that we have not yet proved anything about the dataflow analyses used to
find the rewrites -- we have only shown that the rewrites themselves are
semantically justified.
